ODGI: understanding pangenome graphs

### 2021-10-19 manuscript/main.tex
drafting odgi viz and its placeholder image

%ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
%It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:

@@

% build
We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA. Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes. Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).

@@

% andrea (viz) & erik (draw)

@@

%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.

Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
\textit{odgi viz} and \textit{odgi draw} provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.

\textit{odgi viz} offers a binned, linearized rendering in 1 dimension (1D).

% Explain odgi viz's visualization

However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes, making it easy to grasp the properties of the graph and its complexity.

% Explain odgi viz's color by path position
% Explain odgi viz's color by strandness
% Explain odgi viz's color by coverage

\textit{odgi draw} extends the visualization in 2 dimensions (2D), taking in input the layouts built by \textit{odgi layout}.

% Explain odgi draw's visualization

Figure here with 1D + 2D vizs.

@@

% simon
% HTT exon1
% Only chr8 longs if I can't produce someting nice

@@

% Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.

@@

>>>>

The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:

@@

We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA. Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes. Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).

@@

@@

Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation. We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating pictures of the high-level structure of the graph.

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D). This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
% Explain here odgi viz

However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
% Explain here 3 odgi viz modes

\textit{odgi draw} extends the visualization in 2 dimensions (2D) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}). A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.

@@

% HTT exon1

@@

@@

\input{fig_odgi_viz}
